Making the Best of a Bad Situation

### 2022-03-20 paper.tex
cut in half

%%
%% This is file `sample-authordraft.tex',
%% generated with the docstrip utility.
%%
%% The original source files were:
%%
%% samples.dtx  (with options: `authordraft')
%% 
%% IMPORTANT NOTICE:
%% 
%% For the copyright see the source file.
%% 
%% Any modified versions of this file must be renamed
%% with new filenames distinct from sample-authordraft.tex.
%% 
%% For distribution of the original source see the terms
%% for copying and modification in the file samples.dtx.
%% 
%% This generated file may be distributed as long as the
%% original source files, as listed above, are part of the
%% same distribution. (The sources need not necessarily be
%% in the same archive or directory.)
%%
%% The first command in your LaTeX source must be the \documentclass command.
\documentclass[sigconf,nonacm]{acmart}

@@

@@

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{Making the Best of a Bad Situation}
\subtitle{Picking Random Generators with Multi-Armed Bandits}

@@

\section{Motivation}
% Deductive Program Verifiers like Dafny and Frama-C have proven themselves both
% useful and powerful tools in the fight against un-verified software.  However,
% their adoption is hindered by the fact that graduate-level expertise in
% verification is required in order to verify any non-trivial program. One
% approach to lowering the barrier to entry for deductive verifiers is to
% incorporate lighter-weight formal methods into the mix. While SMT based
% verification is very powerful, its nuances lead to much of the difficulty in
% using verifiers based on it, and so selectively substituting this verification
% technique with a less precise but easier-to-use formal method would lower the
% difficulty while retaining the same workflow and some of the theoretical
% guarantees.

In this extended abstract, we propose and investigate the use of \textit{Property-Based Testing} in
place of SMT solving as a core for program verification tools like Dafny and Frama-C.
Property-Based Testing (PBT) \cite{qc} is a lightweight formal method which checks the
specifications for a program by randomly generating thousands of
inputs, and ensuring that the program meets its specification at each concrete value.

A PBT-based approach to program verification has significant usability benefits
over SMT-based verification. While using SMT-based verifiers, it is often difficult
to interpret the SMT output. When a solver fails to verify a proposition,
either (a) the specification is incorrect, or
(b) the solver was simply unable to find a proof. These two results are often
indistinguishable, and so the only course of action is
to attempt to provide more hints to the solver. With PBT, the only failure mode is through
counter-examples, and so no interpretation is required \footnote{
The counter-examples in a PBT approach are also more easily interpretable. With SMT,
counter-models are encoded in an SMT data format and are not easily lifted
back to program values.
}. PBT can also aid in
the development of SMT-verified programs by allowing the user to test
specifications to gain confidence that they are true before spending costly time
attempting to develop tricky loop invariants.

Moreover, PBT is fully modular, and so it can be used selectively:
randomly-tested code and SMT-verified code can coexist in the same file,
enabling users to take a \textit{gradual verification} approach by gradually
replacing tested code with verified code.

In this paper, we develop a simple deductive program verifier
which is backed by PBT instead of SMT. In \autoref{sec:pbt}, we show how PBT can
be used in place of an SMT solver to do deductive verification. This process is
conceptually simple, but we will see that it hides a major challenge: generating
thousands of independent random input sets which satisfy a function's
precondition is NP-hard. In fact, tackling a small variant of problem is the
bulk of the content of this report, and its main technical contribution. We give
a sketch of this contribution in this section. In brief, our technique first
infers a set of ``candidate" example-generators from the precondition in
question, and then uses a reinforcement-learning-based online learning algorithm
to find the best generator among the set.

In \autoref{sec:luck}, we introduce a subset of a domain-specific language
called \texttt{ALuck} based on \texttt{Luck} \cite{luck}, whose programs are
generators. We discuss the semantics of these generators, and give some examples
of good and bad generators.

In \autoref{sec:sci}, we present our algorithm for inferring candidate
generators for a precondition. In \autoref{sec:bandits}, we discuss the online
learning technique we employ to discover the best generator among our
candidates. Finally, in \autoref{sec:impl}, we discuss our implementation of
this algorithm, present benchmarks, and discuss future work.

\section{PBT, the Generator Problem, and our Approach}
\label{sec:pbt}
At the core of all deductive program verifiers is a decision procedure for Hoare
triples $\{\phi\}\, C \, \{\psi\}$. Such a triple is valid when for all heaps
$\sigma$ satisfying $\phi$, we have that $\psi$ holds of the heap which results
after $C$ is run on $\sigma$. In symbols:

\[
\forall \sigma. \phi(\sigma) \implies \psi(C(\sigma)) \tag{$\ast$}
\]

With traditional SMT-based verifiers, this is decided by compting the weakest
precondition of $\psi$ with respect to $C$, $\texttt{wp}(C,\psi)$ and using SMT
to prove that $\phi$ implies it. With property-based random testing, we can skip
the weakest precondition computation and directly test ($\ast$). PBT tests
formulae like $\forall x. P(x)$ by randomly generating values $a$ (of a
specified type), and then evaluating the concrete boolean expression $P(a)$.
When the property $P$ is an implication $Q \implies R$, as is the case for
($\ast$), this is not very effective. Intuitively, we'd like to test that $R$
holds for a bunch of examples satisfying $Q$, but what actually ends up
happening is that the vast majority of our random examples satisfy $Q \implies
R$ simply because they don't satisfy $Q(x)$. To fix this, PBT relies on
\textit{user-written} generators which generate values \textit{satisfying $Q$},
and then checking that $R$ holds.

While 
In terms of ($\ast$), this means that a user of a verifier backed by PBT would
have to provide programs which generate random heaps $\sigma$ to satisfy the
preconditions of every function they plan to test. This is labor-intensive,
error prone, and diminishes the benefits of PBT over SMT-based verification. To
make the former practical, we need to \textit{automatically} create generators
from the ``source code" of a specification. Unfortunately, this problem is
NP-Hard in even simple cases. If we limit ourselves to preconditions $\phi$
which consist only of arithmetic constraints (which we will in fact do), the
problem of generating hundreds of sets of inputs which satisfy $\phi$ is as hard
as generating \textit{one} set of inputs, which is equivalent to SMT solving.

In the rest of this report, we will focus in on this generation problem, and
present a solution to a specific instance of it. The main cause of the generator
problem described above is that propositions are \textit{declarative}: they
simply state what must hold in order for them to be satisfied, and provide no
instructions on how to generate values which satisfy them. In classical PBT, the
user is required to write \textit{generators} which are programs that emit
random values satisfying a property. Ideally, we would like to automatically
translate declarative propositions into generators, but as stated earlier, this
is arbitrarily hard. Instead, the core of our approach is a translation of
propositions into a large set of \textit{potentially failing} generators. By
relaxing the problem from inferring one \textit{perfect} generator to a few
carefully-chosen generators of unknown (but hopefully high) quality, we make the
problem tractable. Empirically, it seems that among a large enough set of these
automatically-created generator candidates, there is often one which
approximates a generator which a human would have written for the same
proposition. With this in mind, we can then run an online learning algorithm to
learn which among the candidates is the best, while simultaneously generating a
sequence of random inputs for testing purposes.

@@

Before we discuss the generator learning algorithm, we must discuss the class of
propositions for which we will be learning generators, as well as the class of
generators to be learned.

The syntax of preconditions that we plan to handle are shown in
\autoref{fig:precond-syntax}. These conditions are conjunctions and negations of
arithmetical equalities and inequalities with variables ranging over the
function's arguments. The allowable expressions in these inequalities are
essentially multi-linear functions, where each variable occurs with degree at
most one. While this form is restrictive, we have found empirically that the
numerical parts most preconditions of normal functions fall within this class.
This class of preconditions does not represent the upper limit of what our
technique can handle, and we believe that it can be extended to handle
constraints with other numerical operators (division, mod). Moreover, there
appears to be no inherent difficulty to extending the technique to handle
constraints over structured types like lists: we present the algorithm as-is to
simplify the discussion.

\begin{figure}
\caption{Syntax of Preconditions}
\label{fig:precond-syntax}
$$
\begin{array}{llll}
\text{Expressions} & e & ::= & x \; | \; e + e \; | \; -e \;|\; e*e\\
\text{Relations} & R & ::= & = \;|\; \leq \;|\; <\\
\text{Propositions} & P & ::= & P \wedge P \;|\; e R e\\
\end{array}
$$
\end{figure}

@@

More concretely, generators in \texttt{ALuck} are sequences of ``concretize"
operations, written $!x$, and ``constrain" operations, written simply as the
constraint to be added. The syntax of \texttt{ALuck} is shown in
\autoref{fig:aluck-syntax}. These sequences are then evaluated from left to
right while maintaining a pair of mappings: one from concretized variables to
their values, and the other from yet-unconcretized variables to the set of
constraints that have accumulated on them.  When a ``constrain" operation $c$ is
encountered, the constraint $c$ is added to the constraint sets of all of the
variables it mentions. If a constraint mentions no variables, it is checked for
valdity: if the constraint is not valid, the generator fails. When a
``concretize" operation $!x$ is encountered, a value is randomly sampled from
the uniform distribution on the set of possible values\footnote{ Because
integers are bounded by machine word lengths, the ``uniform distribution" does
make sense here, even for unconstrained variables.  } denoted by the constraints
on $x$. This semantics is shown in \autoref{fig:aluck-semantics}, where the
judgment $V,C \vdash s \Downarrow V',C'$ means that the script $s$ evaluates
under concretized-variable-map $V$ and constraint map $C$, and returns updated
variable and constraint maps $V'$ and $C'$.

@@

To give some intuition about how these generator scripts work, we consider
running the example script $s = [0 \leq x, !x, x \leq y, !y]$, which is a
generator script for $P(x,y) = 0\leq x \leq y$. When we run $s$, the constraint
$0 \leq x$ is added to $x$. Next, $x$ is concretized: a value $a$ is sampled
from $\mathcal{U}\left[0,2^{64}-1\right]$. Then, the constraint $a \leq y$ is
added to $y$. Finally, a value $b$ is sampled from
$\mathcal{U}\left[a,2^{64}-1\right]$, and the result $\{x \mapsto a, y \mapsto
b\}$ is returned.

It is crucial to note that generators can fail. For example, the generator
$[!x,!z, x \leq y \leq z, !y]$ will fail if the generated $x$ and $z$ are such
that $z < x$:  the sample domain for $y$ will be empty. Some generators like
this one fail much all the time, and some such as the generator $s$ above never
fail.

\begin{definition}[$p$-Goodness]
We say that a generator script $s$ is $p$-good if it succeeds with probability at least $p$.
\end{definition}

@@

\section{Generator Candidate Inference}
\label{sec:sci}
The first step in our algorithm is to infer a set of generators from a given predicate. We begin by noting that every ordering of the variables in a property immediately determines a well-concretized generator: this procedure is shown below in \autoref{alg:fixed-ordering}. In essence, the procedure works by placing all of the constraints that could possibly appear before a concretization immediately before it.

Because of this, to infer good generators for a property $P$, it suffices to generate good orderings of its variables. Unfortunately, for a property $P$ with $n$ free variables, there are $n!$ such orderings: a very large search space. Fortunately, we can prune this search space by only looking for ``relevant" orderings. To illustrate, consider the property $x \leq y \wedge u \leq v$. There are $24$ different variable orderings on $4$ variables, but there are really only four useful concretization orders: the two possible orders of $x$ and $y$, combined with the two possible orders of $u$ and $v$. Because there are no interactions between $x$ and $u$ or $y$ and $v$, the two orderings $x,u,y,v$ and $x,y,u,v$ will give the same generator. This quotienting allows us to shrink the space of orderings significantly.

To operationalize this, we create an undirected graph from a proposition to encode the important inter-relationships that a concretization order should capture.

\begin{definition}[Proposition Graph]
For a proposition $P$, define $G(P)$ to be the graph whose nodes are variables, with an edge $(x,y)$ when $x$ and $y$ both occur in one of the conjuncts of $P$.
\end{definition}

Then, to generate a concretization ordering, we randomly depth-first search $G(P)$, and list variables in the order that they're visited in the graph.
% Pseudocode for this is shown in \autoref{alg:generate-ordering}.

@@

To generate our set of generator candidates, we repeatedly run this function. This may give repeated generators, and so we filter the result for uniqueness. It's worth noting that getting repeated results is good: it means that we have successfully pruned the search space sufficiently to get a small number of possible generators. The number of generators in our set requires a careful balance. Too few and the set may not contain a generator which succeeds often enough to rapidly generate our desired number of unique inputs. Too many and the learning algorithm will converge too slowly to the best generator in the set. Empirically, we have found that $n^2$ (where $n$ is the number of variables in $P$) is a good number of generators to take.

\section{Generator Learning with Bandits Algorithms}
\label{sec:bandits}
With our bag of generators in hand, we now need to find the best one. The approach we take will be inspired by the Multi-Armed Bandits \cite{gittins1979bandit} problem \footnote{More precisely, Stochastic Bernoulli Bandits} from the theory of reinforcement learning. In short, the multi armed bandits problem describes a situation where an algorithm is repeatedly presented a fixed set of choices. Each choice gives a different (random) reward, and the goal of the game is to maximize the total reward by learning which choice or set of choices gives the best rewards. In our setting, the ``choices" are our generator candidates, and the rewards are given by success or failure of a generator to yield a value. Under this analogy, an algorithm for the Multi-Armed Bandits problem will let us learn which generators give the best results \textit{while simultaneously} generating a stream of valid inputs for the function.

To make the discussion more concrete, the Multi-Armed Bandits problem is described as the following repeated game: at each round $t$, the player plays an action $a_t \in [K]$, and receives a reward $X_{{a_t},t}$, which is a $\{0,1\}$-valued random variable. The random variables $X_{i,t}$ are IID for fixed $i$, and independent for fixed $t$. We write the mean of the $i$th reward variable (for all $t$) $\mu_i$. The goal of the game is to maximize one's reward, and so the goal of a bandits learning algorithm is to learn an adaptive policy, which takes a history of play up until state $t$ (all actions $a_{t'}$ and received rewards $X_{a_{t'},t'}$ for $t' < t$), and produces a new action $a_t$. The metric by which we compare bandits algorthms is \textit{regret}: how much worse they do than the best policy in hindsight.

\begin{definition}[Regret]
Define $i_\star = \argmax_i \mu_i$, and write $\mu_\star = \mu_{i_\star}$. The regret $R(A)$ of an algorithm $A$ over $T$ rounds is defined as
$$
R(A) = T\mu_\star - \mathbb{E}\left[\sum_{t=1}^T X_{A(t),t}\right]
$$
\end{definition}

The algorithm we will use is called UCB1 \cite{auer2002finite}. UCB1 achieves a regret bound of $O(\sqrt{KT\log K})$. While algorithms are known which achieve $O(\sqrt{KT})$ regret, UCB1 is simple to implement and understand, and performs more than well enough for our setting. In \autoref{alg:ucb1}, we present the UCB1 algorithm, modified for use in property-based testing. The algorithm takes generator candidates $g_1,\dots g_k$, the property $P$, a number of rounds to run for $T$, and returns a list of generated values satisfying $P$.

\begin{theorem}
Fix a property $P$ and generators $g_1,\dots,g_K$ which are $p_i$-good. Then,

$$
\lim_{T \to \infty} \mathbb{E}\left[\frac{1}{T}\left|\texttt{ucb1}(g_1,\dots,g_K,P,T)\right|\right] \geq \max_i p_i
$$
\end{theorem}
Intuitively, this theorem justifies our intuition that the ucb1 algorithm ``finds the best generator". As the number of rounds gets large, the ratio of the number of valid inputs generated by \texttt{ucb1} to the number of rounds is no worse than the success probability of the best generator. One can also think of this theorem as saying that, with high probability\footnote{One could do an analysis using Chebyshev's inequality to bound the difference between the empirical frequency of the sequence and its mean}, that the stream of values emitted by an ``infinite run" of \texttt{ucb1} acts like a $(\max_i p_i)$-good generator for $P$.
\begin{proof}
We begin by noting that $r_t = 1$ in an iteration if and only if an element is added to the output list in stage $t$. Therefore, the length of the output $\left|\texttt{ucb1}(g_1,\dots,g_K,P,T)\right|$ is precisely the sum of the ($\{0,1\}$-valued) $r_t$, $\sum_{t=1}^T r_t$. In the notation of the bandit problem, $r_t$ is the (revealed) value of the random variable $X_{A(t),t}$, and so

@@

>>>>

%%
%% This is file `sample-authordraft.tex',
%% generated with the docstrip utility.
%%
%% The original source files were:
%%
%% samples.dtx  (with options: `authordraft')
%% 
%% IMPORTANT NOTICE:
%% 
%% For the copyright see the source file.
%% 
%% Any modified versions of this file must be renamed
%% with new filenames distinct from sample-authordraft.tex.
%% 
%% For distribution of the original source see the terms
%% for copying and modification in the file samples.dtx.
%% 
%% This generated file may be distributed as long as the
%% original source files, as listed above, are part of the
%% same distribution. (The sources need not necessarily be
%% in the same archive or directory.)
%%
%% The first command in your LaTeX source must be the \documentclass command.
\documentclass[sigconf,nonacm,review,anonymous]{acmart}

@@

\usepackage{thmtools} 
\usepackage{thm-restate}

\declaretheorem[name=Theorem,numberwithin=section]{thm}

@@

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{Making the Best of a Bad Situation}
\subtitle{Inferring Random Generators for Numerical Properties with Multi-Armed Bandits}

@@

\section{Motivation}
Property-Based Testing (PBT) \cite{qc} is a program testing method which tests
logical properties of functions by randomly generating thousands of inputs, and ensuring that
the the properties hold at each concrete value. While originally concieved as a more
principled form of unit testing, PBT also works as a lightweight formal verification method
wherein the specifications for a program are \emph{tested} instead of proved.

The main idea behind this approach to verification is that PBT trades off the
full certainty provided by SMT and theorem proving approaches for a significantly streamlined
user experience requiring considerably less user input. To ``verify'' a universally quantified property
$\forall x. P(x) \implies Q(x)$, we generate values satisfying $P$, and check that they satisfy $Q$.
Unfortunately PBT does still requre \emph{some} user input. Users must occasionally write \emph{generators}: programs which
emit random data that satisfies the premise $P$.

Much of the time, this burden can be mitigated by automation, such as when the data structure being generated
is composed entirely of structures that already have generators, or $P$ is described as an inductive
relation \cite{gggir}. However, the case of inferring generators for \emph{numerical} properties is not well
handled by prior work. Numerical properties---those involving constriants over numbers---appear regularly in
systems verification tasks \cite{???}, where structured data described by inductive invariants is rare, inputs regularly
take the form of pointers and integers, and many function preconditions amount to bounds-checking.

In this paper, we develop a method for inferring random generators for a small class of numerical properties akin
to the kind commonly found in systems verification. Because manual generator-writing for numerical properties is especially
tedious, our goal is to infer generators that are comparable to ones that a user might write as a first cut.
The method proceeds by developing a number of generator ``candidates'', expressed in a DSL we call \texttt{ALuck}.
These candidates mimic the way that a human might write a generator for numerical preconditions: by randomly choosing values
one variable at a time, constraining subsequent choices by the previous values chosen. As we will see, it is sometimes the
case that \emph{none} of the candidates are all that good. In this case, we must make the best of a bad situation, and
discover the best candidate among the set.

In \autoref{sec:luck}, we introduce the syntax of the properties we will handle,
as well as the syntax and semantics of \texttt{ALuck}. Next, in \autoref{sec:sci},
we present our algorithm for inferring candidate
generators for a precondition. Then in \autoref{sec:bandits}, we discuss the online
learning technique we employ to discover the best generator among our
candidates. Finally, in \autoref{sec:impl}, we describe our implementation of
this algorithm, present benchmarks, and briefly discuss future work.

% JOE:
% We want to be able to do this so we could do PBT for systems-y code!

% Numerical testing is a really bad situation because we can't really leverage any type
% structure to do the work, it's just *constraints*. Even worse, figuring out a way to
% manually generate these from the constraints is very not fun. (Which goes first? etc etc)
% So we do our best to find some candidates and then pick the best. It's not sexy, but it works.

% In this paper, we develop a simple deductive program verifier
% which is backed by PBT instead of SMT. In \autoref{sec:pbt}, we show how PBT can
% be used in place of an SMT solver to do deductive verification. This process is
% conceptually simple, but we will see that it hides a major challenge: generating
% thousands of independent random input sets which satisfy a function's
% precondition is NP-hard. In fact, tackling a small variant of problem is the
% bulk of the content of this report, and its main technical contribution. We give
% a sketch of this contribution in this section. In brief, our technique first
% infers a set of ``candidate" example-generators from the precondition in
% question, and then uses a reinforcement-learning-based online learning algorithm
% to find the best generator among the set.

% \section{PBT, the Generator Problem, and our Approach}
% \label{sec:pbt}
% At the core of all deductive program verifiers is a decision procedure for Hoare
% triples $\{\phi\}\, C \, \{\psi\}$. Such a triple is valid when for all heaps
% $\sigma$ satisfying $\phi$, we have that $\psi$ holds of the heap which results
% after $C$ is run on $\sigma$. In symbols:

% \[
% \forall \sigma. \phi(\sigma) \implies \psi(C(\sigma)) \tag{$\ast$}
% \]

% With traditional SMT-based verifiers, this is decided by compting the weakest
% precondition of $\psi$ with respect to $C$, $\texttt{wp}(C,\psi)$ and using SMT
% to prove that $\phi$ implies it. With property-based random testing, we can skip
% the weakest precondition computation and directly test ($\ast$). PBT tests
% formulae like $\forall x. P(x)$ by randomly generating values $a$ (of a
% specified type), and then evaluating the concrete boolean expression $P(a)$.
% When the property $P$ is an implication $Q \implies R$, as is the case for
% ($\ast$), this is not very effective. Intuitively, we'd like to test that $R$
% holds for a bunch of examples satisfying $Q$, but what actually ends up
% happening is that the vast majority of our random examples satisfy $Q \implies
% R$ simply because they don't satisfy $Q(x)$. To fix this, PBT relies on
% \textit{user-written} generators which generate values \textit{satisfying $Q$},
% and then checking that $R$ holds.

% While 
% In terms of ($\ast$), this means that a user of a verifier backed by PBT would
% have to provide programs which generate random heaps $\sigma$ to satisfy the
% preconditions of every function they plan to test. This is labor-intensive,
% error prone, and diminishes the benefits of PBT over SMT-based verification. To
% make the former practical, we need to \textit{automatically} create generators
% from the ``source code" of a specification. Unfortunately, this problem is
% NP-Hard in even simple cases. If we limit ourselves to preconditions $\phi$
% which consist only of arithmetic constraints (which we will in fact do), the
% problem of generating hundreds of sets of inputs which satisfy $\phi$ is as hard
% as generating \textit{one} set of inputs, which is equivalent to SMT solving.

% In the rest of this report, we will focus in on this generation problem, and
% present a solution to a specific instance of it. The main cause of the generator
% problem described above is that propositions are \textit{declarative}: they
% simply state what must hold in order for them to be satisfied, and provide no
% instructions on how to generate values which satisfy them. In classical PBT, the
% user is required to write \textit{generators} which are programs that emit
% random values satisfying a property. Ideally, we would like to automatically
% translate declarative propositions into generators, but as stated earlier, this
% is arbitrarily hard. Instead, the core of our approach is a translation of
% propositions into a large set of \textit{potentially failing} generators. By
% relaxing the problem from inferring one \textit{perfect} generator to a few
% carefully-chosen generators of unknown (but hopefully high) quality, we make the
% problem tractable. Empirically, it seems that among a large enough set of these
% automatically-created generator candidates, there is often one which
% approximates a generator which a human would have written for the same
% proposition. With this in mind, we can then run an online learning algorithm to
% learn which among the candidates is the best, while simultaneously generating a
% sequence of random inputs for testing purposes.

@@

The syntax of propositions that we handle are shown in
\autoref{fig:precond-syntax}. These are conjunctions and negations of
arithmetical equalities and inequalities with variables, which range over
the arguments of the function under test. The allowable expressions in these inequalities are
essentially multi-linear functions, where each variable occurs with degree at
most one. While this form is restrictive, we have found empirically that this covers
a wide range of preconditions in systems verification. Moreover, there
appears to be no inherent difficulty to extending the technique to handle
other numerical operators (div, mod), and constraints over structured types like lists: we present the algorithm as-is to
simplify the discussion.

\begin{figure}
\caption{Syntax of Propositions}
\label{fig:precond-syntax}
$$
\begin{array}{llll}
\text{Expressions} & e & ::= & x \; | \; e + e \; | \; -e \;|\; e*e\\
\text{Relations} & R & ::= & = \;|\; \leq \;|\; <\\
\text{Propositions} & P & ::= & P \wedge P \;|\; e R e\\
\end{array}
$$
\end{figure}

@@

More concretely, generators in \texttt{ALuck} are sequences of ``concretize"
operations, written $!x$, and ``constrain" operations, written simply as the
constraint to be added. The syntax of \texttt{ALuck} is shown in
\autoref{fig:aluck-syntax}. These sequences are then evaluated from left to
right while maintaining a pair of mappings: one from concretized variables to
their values, and the other from yet-unconcretized variables to the set of
constraints that have accumulated on them.  When a ``constrain" operation $c$ is
encountered, the constraint $c$ is added to the constraint sets of all of the
variables it mentions. If a constraint mentions no variables, it is checked for
valdity: if the constraint is not valid, the generator \emph{fails}.

When a ``concretize" operation $!x$ is encountered, a value is randomly sampled from
the uniform distribution on the set of possible values\footnote{ Because
integers are bounded by machine word lengths, the ``uniform distribution" does
make sense here, even for unconstrained variables.  } denoted by the constraints
on $x$. This semantics is shown in \autoref{fig:aluck-semantics}, where the
judgment $V,C \vdash s \Downarrow V',C'$ means that the script $s$ evaluates
under concretized-variable-map $V$ and constraint map $C$, and returns updated
variable and constraint maps $V'$ and $C'$.

@@

% To give some intuition about how these generator scripts work, we consider
% running the example script $s = [0 \leq x, !x, x \leq y, !y]$, which is a
% generator script for $P(x,y) = 0\leq x \leq y$. When we run $s$, the constraint
% $0 \leq x$ is added to $x$. Next, $x$ is concretized: a value $a$ is sampled
% from $\mathcal{U}\left[0,2^{64}-1\right]$. Then, the constraint $a \leq y$ is
% added to $y$. Finally, a value $b$ is sampled from
% $\mathcal{U}\left[a,2^{64}-1\right]$, and the result $\{x \mapsto a, y \mapsto
% b\}$ is returned.

% It is crucial to note that generators can fail. For example, the generator
% $[!x,!z, x \leq y \leq z, !y]$ will fail if the generated $x$ and $z$ are such
% that $z < x$:  the sample domain for $y$ will be empty. Some generators like
% this one fail much all the time, and some such as the generator $s$ above never
% fail.

The fact that generators can fail is crucial: different generators for the same property can fail more or
less often, and we would like to infer generators which fail infrequently.
\begin{definition}[$p$-Goodness]
We say that a generator script $s$ is $p$-good if it succeeds with probability at least $p$.
\end{definition}

@@

\section{Generator Candidate Inference}
\label{sec:sci}
The first step in our algorithm is to infer a set of generators from a given
predicate. We begin by noting that every ordering of the variables in a property
immediately determines a well-concretized generator:
this procedure is shown in the Appendix in \autoref{alg:fixed-ordering}.
In essence, the procedure works by placing all of the constraints that could possibly
appear before a concretization immediately before it.

Because of this, to infer generators for a property $P$, it suffices to
generate orderings of its variables. Unfortunately, for a property $P$ with
$n$ free variables, there are $n!$ such orderings: a very large search space.
Fortunately, we can prune this search space by only looking for ``relevant"
orderings. If some variables are not related to others, the well-concretization
condition won't care the relative order in which they're generated.

% To illustrate, consider the property $x \leq y \wedge u \leq v$.
% There are $24$ different variable orderings on $4$ variables, but there are
% really only four useful concretization orders: the two possible orders of $x$
% and $y$, combined with the two possible orders of $u$ and $v$. Because there are
% no interactions between $x$ and $u$ or $y$ and $v$, the two orderings $x,u,y,v$
% and $x,y,u,v$ will give the same generator. This quotienting allows us to shrink
% the space of orderings significantly.

To operationalize this insight, we build a graph $G(P)$ whose nodes are variables
with an edge $(x,y)$ when $x$ and $y$ both occur in one of the conjuncts of $P$. In
this graph, ``unrelated'' variables live in different connected components.
Then, to generate a concretization ordering, we depth-first search $G(P)$, randomly
choosing the next neighbor to explore,
and list variables in the order that they're visited in the graph.
Since different connected components are listed separately, unrelated variables
will not 
The fact that related variables lie in separate connected components means
that 
% we create an undirected graph from a proposition to
% encode the important inter-relationships that a concretization order should
% capture.

% Pseudocode for this is shown in \autoref{alg:generate-ordering}.

@@

To generate our set of generator candidates, we repeatedly run this function.
This may give repeated generators, and so we filter the result for uniqueness.
The number of generators in our set requires a careful
balance. Too few and the set may not contain a generator which succeeds often
enough to rapidly generate our desired number of unique inputs. Too many and the
learning algorithm will converge too slowly to the best generator in the set.
Empirically, we have found that $n^2$ (where $n$ is the number of variables in
$P$) is a number of generators to take.

\section{Generator Learning with Bandits Algorithms}
\label{sec:bandits}
With our bag of generators in hand, we now need to find the best one. The
approach we take will be inspired by the Multi-Armed Bandits
\cite{gittins1979bandit} problem\footnote{More precisely, Stochastic Bernoulli
Bandits} from the theory of reinforcement learning. In short, the multi armed
bandits problem describes a situation where an algorithm is repeatedly presented
a fixed set of choices. Each choice gives a different (random) reward, and the
goal of the game is to maximize the total reward. Solving this optimization problem
online is too difficult in an adversarial setting, so algorithms for the multi-armed
bandits problem aim to instead minimize \emph{regret}: how much worse their total reward
was than the best possible total reward in hindsight. To achieve this, bandits algorithms
learn which actions have historically given more reward, and play those more frequently.
In our setting, the ``choices" are our
generator candidates, and the rewards are given by success or failure of a
generator to yield a value. Under this analogy, an algorithm for the Multi-Armed
Bandits problem will let us learn which generators give the best results
\textit{while simultaneously} generating a stream of valid inputs for the
function.

% To make the discussion more concrete, the Multi-Armed Bandits problem is
% described as the following repeated game: at each round $t$, the player plays an
% action $a_t \in [K]$, and receives a reward $X_{{a_t},t}$, which is a
% $\{0,1\}$-valued random variable. The random variables $X_{i,t}$ are IID for
% fixed $i$, and independent for fixed $t$. We write the mean of the $i$th reward
% variable (for all $t$) $\mu_i$. The goal of the game is to maximize one's
% reward, and so the goal of a bandits learning algorithm is to learn an adaptive
% policy, which takes a history of play up until state $t$ (all actions $a_{t'}$
% and received rewards $X_{a_{t'},t'}$ for $t' < t$), and produces a new action
% $a_t$. The metric by which we compare bandits algorthms is \textit{regret}: how
% much worse they do than the best policy in hindsight.

% \begin{definition}[Regret]
% Define $i_\star = \argmax_i \mu_i$, and write $\mu_\star = \mu_{i_\star}$. The regret $R(A)$ of an algorithm $A$ over $T$ rounds is defined as
% $$
% R(A) = T\mu_\star - \mathbb{E}\left[\sum_{t=1}^T X_{A(t),t}\right]
% $$
% \end{definition}

The algorithm we will use is called UCB1 \cite{auer2002finite}. While more sophisticated
bandits algorithms exist, this one is sufficient for our purposes. When given a list of generators
$g_1,\dots g_k$ and a number of rounds $T$ to run for, UCB1 emits some number $m \leq T$ of values generated by the
generators it was given. The main upshot is that UCB1 rapidly learns which generator among the ones it was given
fails the \emph{least often}, and so the ratio success ratio $\frac{m}{T}$ of this derived generator should, in the limit,
be no worse than the best generator.

% UCB1 achieves a
% regret bound of $O(\sqrt{KT\log K})$. While algorithms are known which achieve
% $O(\sqrt{KT})$ regret, UCB1 is simple to implement and understand, and performs
% more than well enough for our setting. In \autoref{alg:ucb1}, we present the
% UCB1 algorithm, modified for use in property-based testing. The algorithm takes
% generator candidates $g_1,\dots g_k$, the property $P$, a number of rounds to
% run for $T$, and returns a list of generated values satisfying $P$.

\begin{restatable}{thm}{regretthm}
Fix a property $P$ and generators $g_1,\dots,g_K$ which are $p_i$-good. Then,

$$
\lim_{T \to \infty} \mathbb{E}\left[\frac{1}{T}\left|\texttt{ucb1}(g_1,\dots,g_K,P,T)\right|\right] \geq \max_i p_i
$$
\end{restatable}

% Intuitively, this theorem justifies our intuition that the ucb1 algorithm
% ``finds the best generator". As the number of rounds gets large, the ratio of
% the number of valid inputs generated by \texttt{ucb1} to the number of rounds is
% no worse than the success probability of the best generator. One can also think
% of this theorem as saying that, with high probability\footnote{One could do an
% analysis using Chebyshev's inequality to bound the difference between the
% empirical frequency of the sequence and its mean.}, that the stream of values
% emitted by an ``infinite run" of \texttt{ucb1} acts like a $(\max_i p_i)$-good
% generator for $P$.

@@

\newpage
\appendix
\section*{Appendix}

\regretthm*
\begin{proof}
We begin by noting that $r_t = 1$ in an iteration if and only if an element is added to the output list in stage $t$. Therefore, the length of the output $\left|\texttt{ucb1}(g_1,\dots,g_K,P,T)\right|$ is precisely the sum of the ($\{0,1\}$-valued) $r_t$, $\sum_{t=1}^T r_t$. In the notation of the bandit problem, $r_t$ is the (revealed) value of the random variable $X_{A(t),t}$, and so
